---
title: "Untitled"
format: html
editor: visual
---

## Quarto

```{r}
library(dplyr)
library(stringr)
library(naniar) 


dataset_freezed2Nov25 <- read.csv("../data/dataset_freezed2Nov25.csv")

dataset_freezed2Nov25 <- dataset_freezed2Nov25 %>% 
  mutate(age_cat = case_when(
      age <= 18 ~ "Child",
      age < 60 ~  "Adult",
      TRUE ~ "Elderly" 
    ),
    index_child = case_when(
      index_child == 1 ~ "Index case age < 18 years",
      index_child == 0 ~ "Index case age > 18 years"
    )
    # ,mask_ind = case_when(
    #   mask_ind == 1 ~ "Yes, n(%)",
    #   mask_ind == 0 ~ "No, n(%)"
    # ),
    # mask_mem = case_when(
    #   mask_mem == 1 ~ "Yes, n(%)",
    #   mask_mem == 0 ~ "No, n(%)"
    # ),
    # sepbed = case_when(
    #   sepbed == 1 ~ "Yes, n(%)",
    #   sepbed == 0 ~ "No, n(%)"
    # ),
     # Number of bedrooms
    # bedroom_cat = case_when(
    #   bedroom_num == "1" ~ "1 room, n(%)",
    #   bedroom_num == "2" ~ "2 rooms, n(%)",
    #   bedroom_num == "3" ~ "3 rooms, n(%)",
    #   bedroom_num == "4" ~ "4 rooms, n(%)",
    #   bedroom_num == "5" ~ "5 rooms, n(%)",
    #   bedroom_num == "6" ~ "6 rooms, n(%)",
    #   bedroom_num == "7" ~ "7 rooms, n(%)",
    # ),    
    # Number of bathrooms
    # toiletcat = case_when(
    #   toilet_num == "1" ~ "1 room, n(%)",
    #   toilet_num == "2" ~ "2 rooms, n(%)",
    #   toilet_num == "3" ~ "3 rooms, n(%)",
    #   toilet_num == "4" ~ "4 rooms, n(%)",
    #   toilet_num == "5" ~ "5 rooms, n(%)"
    # )
    ) %>% 
  group_by(hhid) %>%
  # Apply the value to the whole family
  mutate(exclude_mem = exclude_mem[index == 1][1]) %>%
  ungroup()


HH_var <- c("index_child",'household_size',"household_sizemorethan4",'total_inhouse',
                 "include_all_members",'bedroom_num', 'toilet_num',
                 'separate_bed',"wash_sink",'ven_aircon', "openaircond",
                 "havepet","toilet_cat" ,"bedroom_cat", "mask_mem", "include_all_members")
contact_var <- c("sex", "age", "age_cat","edu"," relatetoindex","hazbmi_cat",
"condition_hypertension","  condition_kidney_disease","condition_diabetes",
"smoking","total_vac_cat","vac_cat","latest_vac_cat","period_vac_cat",
"everinf","mask_mem","result_IgG_day1_mod",
                 "result_IgM_day1_mod")

# take off useless variables



```

# Look at missing data in index df

```{r}

# Remove columns with only NA

index_col <- c("pid","hhid","index","index_child","day1_date","sex","age","age_cat","edu","hazbmi_cat","symptom","total_vac_cat","vac_cat","latest_vac_cat","period_vac_cat","mask_ind","sepbed","lineagecat","ngene_day1","ngene_d1_cat","sgene_day1","sgene_d1_cat","include_all_members","household_size","exclude_mem")

index_df <- dataset_freezed2Nov25 %>% 
  filter(index == "1") %>% 
  select(all_of(index_col))

index_df <- index_df %>% 
  mutate(edu = ifelse(index_child=="Index case age < 18 years", "Child", edu))


gg_miss_var(index_df, show_pct = TRUE)

```

Almost no data is missing there !

# Improving estimation of household transmission

1)**Exclude the Households with multiple index cases (15 households): The Multiple chains of transmissions makes it too complex to infer the SAR.**

How can we know if the household had multiple index cases ?

## **2) Imputation of status of individual excluded members**

La il faut regarder pour les membres exclus les caracteristiques du reste pour deviner l'age et le sexe.

Imputation de données manquantes au sein d'une structure hiérarchique (individus nichés dans des foyers)

Let's start with imputation from age and sex.

```{r}

```

### Contruct imputation df

we construct the imputation dataframe with the variables we will use in our model.

```{r}

imputation_df <- dataset_freezed2Nov25 %>%
  group_by(hhid) %>%
  mutate(age_index = age_cat[index == 1],
         n_observed_contacts    = household_size - (exclude_mem +1),
         prop_infected_contacts = sum(result_infected[index == 0] == "Detected",
                                      na.rm = TRUE)/n_observed_contacts,
         n_infected_contacts    = sum(result_infected[index == 0] == "Detected", na.rm = TRUE),
    n_missing_contacts     = exclude_mem,
    # Crowding proxy
    #persons_per_room       = household_size / bedroom_num)
    age_mean   = mean(age, na.rm = TRUE),
    age_min     = min(age, na.rm = TRUE),
    age_max    = max(age, na.rm = TRUE),
    female_prop       = mean(sex == "Female", na.rm = TRUE),
    adult_nb        = sum(age >= 18, na.rm = TRUE),
    child_nb        = sum(age < 18, na.rm = TRUE),
    elderly_nb      = sum(age >= 60, na.rm = TRUE)
  ) %>% 
  select(hhid,age_index,n_observed_contacts, prop_infected_contacts, n_infected_contacts, n_missing_contacts, age_mean, age_min, age_max, adult_nb, child_nb,elderly_nb) %>% 
  distinct()


  
```

## Create the df for included members

```{r}
included_members_df <- df %>%
  select(hhid, index, age_cat, result_infected, household_size, exclude_mem) %>% 
  mutate(statut = "included")

included_members_df <- left_join(included_members_df, imputation_df, by = "hhid")

```

## Create the rows for the excluded members

```{r}
# Identify households with excluded members
hh_excluded <- dataset_freezed2Nov25 %>%
  group_by(hhid) %>%
  slice(1) %>%                          # one row per household
  ungroup() %>%
  filter(exclude_mem > 0) %>%
  select(hhid, exclude_mem,
         household_size)                   # adapt to your actual column names

# Create one new row per excluded member
new_rows <- hh_excluded %>%
  rowwise() %>%
  reframe(
    hhid            = hhid,
    pid             = seq_len(exclude_mem), # temporary IDs
    index = 0,
    age_cat       = NA_character_,
    result_infected       = NA_character_,
    household_size    = household_size,
    exclude_mem = exclude_mem,
    statut          = "missing",
  ) %>% select(-pid)

excluded_members_df <- left_join(new_rows, imputation_df, by = "hhid")

```

```{r}
full_data <- rbind(included_members_df, excluded_members_df)
```

### Train the data on only complete household

```{r}
complete_HH <- dataset_freezed2Nov25 %>% 
  select("hhid","include_all_members") %>% 
  filter(include_all_members == "All members included") %>% 
  pull(hhid)

train_data <- full_data %>% 
  filter(hhid %in% complete_HH,
         statut == "included") %>% 
  mutate(age_cat = as.factor(age_cat)) %>% 
  group_by(hhid) %>% 
  mutate(
    child_nb_loo    = child_nb   - (age_cat == "Child"),
    adult_nb_loo    = adult_nb   - (age_cat == "Adult"),
    elderly_nb_loo  = elderly_nb - (age_cat == "Elderly")
  ) %>%
  ungroup()
```

```{r}

age_weights <- c(
  "Adult" = nrow(train_data) / (3 * sum(train_data$age_cat == "Adult")),
  "Child" = nrow(train_data) / (3 * sum(train_data$age_cat == "Child")),
  "Elderly" = nrow(train_data) / (3 * sum(train_data$age_cat == "Elderly"))
)

library(randomForest)

model_age_group <- randomForest(
  age_cat ~ child_nb_loo + adult_nb_loo + elderly_nb_loo  + household_size,
  data      = train_data,
  ntree     = 500,
  classwt = age_weights,
  importance = TRUE
)

```

```{r}
print(model_age_group)
```

Without class weight, we find an error rate of 0.11, which is great news ! We are quite efficient into knowing the age of our missing people.

With class weight it upgrade to 18%, but we nay have a model more suited to the reality.

# Train the data to predict infection on complete HH

Let's prepare the data first

```{r}
train_infection <- full_data %>%
  filter(hhid %in% complete_HH,
         hhid != "1",
         statut == "included",
         index == 0) %>%
  mutate(
  result_infected = ifelse(result_infected == "Detected",1,0),
  result_infected = as.factor(result_infected),
  age_cat = as.factor(age_cat)) %>% 
  group_by(hhid) %>%
  mutate(
    # Exclude current member from infection summaries
    n_infected_loo = n_infected_contacts - (result_infected == "Detected"),
    n_observed_loo = n_observed_contacts - 1,
    prop_infected_loo = if_else(
      n_observed_loo > 0,
      n_infected_loo / n_observed_loo,
      0),
    child_nb_loo    = child_nb   - (age_cat == "Child"),
    adult_nb_loo    = adult_nb   - (age_cat == "Adult"),
    elderly_nb_loo  = elderly_nb - (age_cat == "Elderly"),
    result_infected = as.factor(result_infected)
  ) %>%
  ungroup() 
      

```

Random Forest with class weights

```{r}
library(broom)

# Add a class weight to balance 
class_weights <- c(
  "0" = nrow(train_infection) / (2 * sum(train_infection$result_infected == 0)),
  "1" = nrow(train_infection) / (2 * sum(train_infection$result_infected == 1))
)

model_infection <- randomForest(
  factor(result_infected) ~ age_cat + age_index + n_infected_loo + n_observed_loo +
    prop_infected_loo,
  data   = train_infection,
  ntree = 500,
  classwt    = class_weights,
  importance = TRUE
)

```

```{r}
print(model_infection)
```

Error rate de 10 % juste avec un random forest, en vrai c'est ok, essayons de le rendre plus parcimonieux :

```{r}
model_infection_p <- randomForest(
  factor(result_infected) ~ age_cat  +
    prop_infected_loo,
  data   = train_infection,
  ntree = 500,
  classwt    = class_weights,
  importance = TRUE
)

print(model_infection_p)
```

Ce modele a 9% d'error rate, alors que c'est juste avec 2 parametres !

I need to treat all my missing members jointly

# Sequential iteration

Let's try to apply the sequential iteration

```{r}
# ============================================================
# STEP 2 : Construire les variables CORRIGÉES pour l'imputation
# incluant les manquants dans le dénominateur
# ============================================================

# Pour chaque membre manquant à imputer, les predicteurs
# doivent refléter la composition OBSERVÉE + déjà imputés

impute_age_sequential <- function(hhid_target, data, model_age) {
  
  famille <- data %>% filter(hhid == hhid_target)
  n_manquants <- first(famille$household_size) - nrow(famille)
  print("Nombre d'habitant manquants: ", n_manquants)
  print(n_manquants)
  ages_imputed <- c()
  
  for (vague in 1:n_manquants) {
    # Recalculer les _loo en intégrant les déjà imputés
    child_so_far   <- sum(famille$age_cat == "Child",   na.rm = TRUE) +
                      sum(ages_imputed == "Child")
    adult_so_far   <- sum(famille$age_cat == "Adult",   na.rm = TRUE) +
                      sum(ages_imputed == "Adult")
    elderly_so_far <- sum(famille$age_cat == "Elderly", na.rm = TRUE) +
                      sum(ages_imputed == "Elderly")
    
    # Ligne fictive pour le membre manquant à imputer
    new_member <- data.frame(
      child_nb_loo   = child_so_far,    # exclut le membre cible (LOO)
      adult_nb_loo   = adult_so_far,
      elderly_nb_loo = elderly_so_far,
      household_size = first(famille$household_size)
    )
    
    # Prédire l'âge
    age_pred <- predict(model_age, newdata = new_member)
    ages_imputed <- c(ages_imputed, as.character(age_pred))
    
    cat(sprintf("Famille %s | Vague %d : age imputé = %s\n",
                hhid_target, vague, age_pred))
  }
  
  return(ages_imputed)
}

# ============================================================
# STEP 3 : Appliquer à toutes les 57 familles
# ============================================================


```

```{r}

#impute_age_sequential(8, dataset_freezed2Nov25, model_age = model_age_group)


results_imputation <- lapply(famille_summary$hhid, function(id) {
  ages <- impute_age_sequential(id, dataset_freezed2Nov25, model_age_group)
  data.frame(hhid = id, age_cat = ages, 
             vague = seq_along(ages))
}) %>% bind_rows()

print(results_imputation)
```

### Let's try to plot the distribution to see if it makes sense

```{r}
library(ggplot2)

observed_age <- dataset_freezed2Nov25 %>% 
  select(hhid, age_cat) %>% 
  mutate(source = "observed")

imputed_age <- results_imputation %>% 
  select(hhid, age_cat) %>% 
  mutate( source = "imputed")

p <- bind_rows(observed_age, imputed_age) %>%
  count(age_cat, source) %>% 
  group_by(source) %>%
  mutate(prop = n / sum(n)) %>%   # proportions pour comparaison équitable
  ggplot(aes(x = age_cat, y = prop, fill = source)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.85) +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)),
            position = position_dodge(width = 0.9),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("observed" = "#457B9D", "imputed" = "#E63946")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = " age_cat distribution : Observed vs Imputed",
    subtitle = paste0("n observed = ", nrow(observed_age),
                      " | n imputed = ", nrow(imputed_age)),
    x        = "Age category",
    y        = "Proportion",
    fill     = "Source"
  ) +
  theme_minimal(base_size = 13)

p
```

My model has tendencies to give more adults, even with weighted class. It would be time to add a household structure if possible to the model .

# TEST

### Essayons avec une loi poisson

On peut essayer les memes stats que Patumrat, a savoir une log link function

```{r}
model_infection_poisson <- glm(
  result_infected ~ age_cat  + n_infected_loo + n_observed_loo +
    prop_infected_loo,
  data   = train_infection,
  family = poisson(link = "log")
)
```

```{r}
summary(model_infection_poisson)
```

```{r}
library(caret)

# Verify
levels(train_infection$result_infected)
# Should be: "No" "Yes"

# Then rerun
model_cv <- train(
  result_infected ~ age_cat + prop_infected_loo,
  data      = train_infection,
  method    = "glm",
  family    = binomial(link = "logit"),
  trControl = ctrl,
  metric    = "ROC"
)



summary(model_cv)
```

# 

# Let's try with mice

```{r}
library("mice")
# Columns to use in the imputation model
cols_for_imputation <- c(
  # Variables to impute
  "age_cat",
  "result_infected",
  
  # Predictors for age_cat (mirrors model_age_group)
  "age_index",
  "household_size",        # OR the nb_ components — not both
  "child_nb_loo",          # remove if keeping household_size
  "adult_nb_loo",
  "elderly_nb_loo",
  
  # Predictors for result_infected (mirrors model_infection_p)
  "prop_infected_loo"
  
)

train_imp <- train_infection[, cols_for_imputation]


# Tell mice: when imputing age_cat, use exactly these predictors
pred_matrix <- make.predictorMatrix(train_imp)

# Zero out everything for age_cat row first
pred_matrix["age_cat", ] <- 0

# Then allow only the same vars your RF used
pred_matrix["age_cat", c("child_nb_loo", "adult_nb_loo", 
                          "elderly_nb_loo", "age_index", 
                          "household_size")] <- 1


# Use your full training data (including result_infected for imputation)
imp <- mice(
  train_imp,
  m               = 20,
  method          = "pmm",
  predictorMatrix = pred_matrix,
  maxit           = 10,
  seed            = 123,
  printFlag       = FALSE
)

complete(imp)
```

```{r}
famille_summary <- full_data %>%
  filter( statut == "included") %>% 
  group_by(hhid) %>%
  summarise(
    n_observed_contacts    = n(),
    n_missing_contacts   = first(n_missing_contacts),
    child_obs     = sum(age_cat == "Child",   na.rm = TRUE),
    adult_obs     = sum(age_cat == "Adult",   na.rm = TRUE),
    elderly_obs   = sum(age_cat == "Elderly", na.rm = TRUE),
    age_index     = first(age_index),
    household_size = first(household_size)
  ) %>%
  filter(n_missing_contacts > 0)

cat("Familles avec membres manquants :", nrow(famille_summary), "\n")
cat("Distribution du nb de manquants :\n")
print(table(famille_summary$n_missing_contacts))
```
