---
title: "Untitled"
format: html
editor: visual
---

# Unicef Mics data

First, we look at the household members data , named hl .

```{r}
library(haven)
hl <- read_sav("../Thailand MICS6 2022 SPSS Datasets/hl.sav")

summary_hl <- look_for(hl)
```

```{r}
library(haven)
library(dplyr)


# Household composition
hl_summary <- hl %>%
  #Select only Chiang Mai families
  filter(HH7A == 50) %>%
  mutate(
    is_child   = HL6 >= 0  & HL6 <= 17,
    is_adult   = HL6 >= 18 & HL6 <= 59,
    is_elderly = HL6 >= 60
  ) %>%
  group_by(HH1,HH2) %>%
  summarise(
    household_size = n(),
    n_child     = sum(is_child,   na.rm = TRUE),
    n_adult       = sum(is_adult,   na.rm = TRUE),
    n_elderly      = sum(is_elderly, na.rm = TRUE),
    .groups = "drop"
  )

summary(hl)

# Weighted frequency for each composition cell


```

### Look at HH variables

let's look at the complete households, we check if it is the same value as computed in hl_summary

```{r}
hh <- read_sav("../Thailand MICS6 2022 SPSS Datasets/hh.sav")

summary_hh <- look_for(hh)

complete_hh <- hh %>%
  #Select only Chiang Mai families
  filter(HH7A == 50) %>%
  # Let's have a look at the household size
  select(HH1,HH2,HH48,hhweight) %>% 
  # DROP the na
  filter(!is.na(HH48))

complete_df <- hl_summary %>% 
  left_join(complete_hh, by=c("HH1","HH2")) %>% 
  mutate(same_value_size = household_size==HH48)
```

The values are consistent, we can then trust them :)

### Adapt the data to our study

We need to have at least one child, at least 2 people in the house.

```{r}
complete_df <- complete_df %>% 
  filter(household_size > 1,
         n_child >0)

comp_table <- complete_df %>%
  group_by(household_size, n_child, n_adult, n_elderly) %>%
  summarise(
    count      = n())
```

# Hot Deck Imputation

### find_donors function

find a potential donor based on household size, and the observed member in the family.

```{r}
find_donors <- function(obs_child, obs_adult, obs_elderly, 
                        hh_size,donor_pool) {
  
  donor_pool %>%
    filter(
      household_size == hh_size,
      n_child   >= obs_child,
      n_adult   >= obs_adult,
      n_elderly >= obs_elderly
    ) %>% 
    mutate(freq = count / sum(count))
}

donors <- find_donors(1,2,1,6,comp_table)

```

### draw_donor function

Out of all the possible donors, pool one based on the freq out each household composition.

```{r}
draw_donor <- function(donors) {
  
  if (nrow(donors) == 0) return(NULL)
  
  # Weighted random draw
  idx <- sample(
    nrow(donors),
    size = 1,
    prob = donors$freq
  )
  
  donors[idx,c("household_size",'n_child','n_adult','n_elderly')]
}

pooled_donor <- draw_donor(donors)

# Imputed missing counts
c(
  child_imputed  = pooled_donor$n_child   - 1,  # subtract observed
  adult_imputed  = pooled_donor$n_adult   - 1,
  elderly_imputed = pooled_donor$n_elderly - 0
)
```

### multiple hot deck function

We pool one donor for each incomplete household

```{r}
multiple_hot_deck <- function(incomplete_HH, donor_pool) {
  
  # Create an empty dataframe
  all_imputation <- data.frame(
    hhid = integer(),
    household_size = integer(),
    n_child = integer(),
    n_adult = integer(),
    n_elderly = integer(),
    imp_child = integer(),
    imp_adult = integer(),
    imp_elderly = integer()
    
  )
  
  for (i in 1:nrow(incomplete_HH)) {
    
    # Find the potential donors for this set:
    donors <- find_donors(
      obs_child   = incomplete_HH$n_child[i],
      obs_adult   = incomplete_HH$n_adult[i],
      obs_elderly = incomplete_HH$n_elderly[i],
      hh_size     = incomplete_HH$household_size[i],
      donor_pool  = donor_pool
    )
    
    # Pool the donor and compute imputation variable 
    pooled_donor <- draw_donor(donors) 
    
    
    # Deal with when there is no pool donor
    if (is.null(pooled_donor)) {
        pooled_donor <- data.frame(
            hhid = incomplete_HH[i,]$hhid,
            household_size = incomplete_HH[i,]$household_size,
            n_child = NA,
            n_adult = NA,
            n_elderly = NA,
            imp_child = NA,
            imp_adult = NA,
            imp_elderly = NA
            
          )
    }
    # Otherwise, we compute what has been imputed 
    else {
          pooled_donor <- pooled_donor %>% 
      mutate(hhid = incomplete_HH[i,]$hhid,
             imp_child = n_child - incomplete_HH[i,]$n_child,
             imp_adult = n_adult - incomplete_HH[i,]$n_adult,
             imp_elderly = n_elderly - incomplete_HH[i,]$n_elderly
             )
      
    }
  
    
  
    all_imputation <- bind_rows(all_imputation,pooled_donor)
  }

  return(all_imputation)
  
  

}

```

## Import the data from the study

we look at all the incomplete hh from our study.

```{r}
library(dplyr)
dataset_freezed2Nov25 <- read.csv("../data/dataset_freezed2Nov25.csv")

all_HH <- dataset_freezed2Nov25 %>% 
  mutate(age_cat = case_when(
      age <= 18 ~ "Child",
      age < 60 ~  "Adult",
      TRUE ~ "Elderly" 
    ),
    ) %>% 
  group_by(hhid) %>%
  # Apply the value to the whole family
  mutate(exclude_mem = exclude_mem[index == 1][1]) %>% 
    group_by(hhid) %>%
  mutate(
    n_missing_contacts     = exclude_mem,
    n_adult        = sum(age >= 18, na.rm = TRUE),
    n_child        = sum(age < 18, na.rm = TRUE),
    n_elderly      = sum(age >= 60, na.rm = TRUE)
  ) %>% 
  select(hhid,household_size, n_adult, n_child, n_elderly, n_missing_contacts)

# Select our useful variables

incomplete_HH <- all_hh %>% 
  filter(n_missing_contacts > 0) %>% 
  distinct()
```

### test on the incomplete data

```{r}
print(incomplete_HH[1,])
mhd <- multiple_hot_deck(incomplete_HH, comp_table)
```

## validation on the complete households data

### function to remove members

first let's have a look at the household size distribution betwwen complete and incomplete hh

```{r}
library(ggplot2)

all_HH %>%
  distinct() %>% 
  mutate(
    group = ifelse(n_missing_contacts >0, "incomplete", "complete")) %>%
  group_by(household_size,group) %>% 
  mutate(count = n()) %>% 
  group_by(group) %>% 
  mutate(prop = count / n()) %>% 
  ggplot(aes(x = household_size, y = prop, fill = group)) +
  geom_col(position = "dodge") +
  labs(title = "Household size distribution: complete vs incomplete")
```

We mainly have complete household of size 3, **The 34 complete household is not a good validation dataset for our problem .**

# Infection imputation

our idea was to use the infection state of other members, aswell as the age group, and possibly the contact matrix(by age).

This is not a standard imputation problem — it is a probabilistic inference problem on a transmission network.

We could :

- create an exposure score to the disease, based on the infection state of the household.

```{r}

```