---
title: "Sensitivity analyses"
subtitle: "VERDI household SARS-CoV-2 transmission model"
format: html
editor: visual
---

# Overview

Systematic comparison of the corrected SAR and posterior of $q$ across all planned
sensitivity scenarios. Each scenario re-runs the same MCMC sampler with one assumption
changed. Results are cached to `results/` so re-rendering is fast after the first run.

**Scenarios covered:**

| ID | Description | What changes |
|---|---|---|
| `primary` | Baseline (τ=2, Prem 2021 matrix) | — |
| `tau1` | Pre-enrollment window τ=1 | `build_hh_list(tau=1)` |
| `tau3` | Pre-enrollment window τ=3 | `build_hh_list(tau=3)` |
| `naive` | Age-naive (no contact matrix) | `C_mat` = all-ones |
| `gamma05` | Density-dependent contacts γ=0.5 | `C_mat` scaled per household |
| `gamma10` | Frequency-dependent contacts γ=1.0 | `C_mat` scaled per household |

------------------------------------------------------------------------

# Setup

```{r}
library(dplyr)
library(ggplot2)
library(haven)

source("R/model_functions.R")

set.seed(2024)

N_ITER   <- 10000
N_BURNIN <- 2000
N_THIN   <- 5
```

------------------------------------------------------------------------

# Shared data (loaded once)

```{r}
# ── Study dataset ────────────────────────────────────────────────────────────
dataset <- read.csv("data/dataset_freezed2Nov25.csv") %>%
  mutate(
    age_cat = factor(
      case_when(age <= 17 ~ "Child", age < 60 ~ "Adult", TRUE ~ "Elderly"),
      levels = c("Child", "Adult", "Elderly")
    ),
    infected_bin = as.integer(result_infected == "Detected")
  ) %>%
  group_by(hhid) %>%
  mutate(exclude_mem = exclude_mem[index == 1][1]) %>%
  ungroup()

n_obs_contacts  <- sum(dataset$index == 0 & !is.na(dataset$infected_bin))
n_obs_infected  <- sum(dataset$index == 0 & dataset$result_infected == "Detected",
                       na.rm = TRUE)
n_total_missing <- sum(dataset$exclude_mem[dataset$index == 1], na.rm = TRUE)

cat("Observed contacts:", n_obs_contacts,
    "| Observed infected:", n_obs_infected,
    "| Total missing:", n_total_missing, "\n")

# ── MICS comp_table (used for all scenarios) ─────────────────────────────────
hl      <- read_sav("Thailand MICS6 2022 SPSS Datasets/hl.sav")
hh_mics <- read_sav("Thailand MICS6 2022 SPSS Datasets/hh.sav")

comp_table <- hl %>%
  filter(HH7A == 50) %>%
  mutate(is_child = HL6 <= 17, is_adult = HL6 >= 18 & HL6 <= 59,
         is_elderly = HL6 >= 60) %>%
  group_by(HH1, HH2) %>%
  summarise(household_size = n(), n_child = sum(is_child),
            n_adult = sum(is_adult), n_elderly = sum(is_elderly),
            .groups = "drop") %>%
  filter(household_size > 1, n_child > 0) %>%
  group_by(household_size, n_child, n_adult, n_elderly) %>%
  summarise(count = n(), .groups = "drop")

# ── Prem 2021 home contact matrix for Thailand ───────────────────────────────
prem_cache <- "data/contact_home.rdata"
if (!file.exists(prem_cache)) {
  download.file(
    paste0("https://raw.githubusercontent.com/kieshaprem/",
           "synthetic-contact-matrices/master/output/",
           "syntheticmatrices/contact_home.rdata"),
    prem_cache, mode = "wb", quiet = TRUE
  )
}
load(prem_cache)

child_bands   <- 1:4; adult_bands <- 5:12; elderly_bands <- 13:16
groups <- list(Child = child_bands, Adult = adult_bands, Elderly = elderly_bands)

C_prem <- matrix(0, 3, 3, dimnames = list(names(groups), names(groups)))
for (a in names(groups))
  for (b in names(groups))
    C_prem[a, b] <- mean(rowSums(contact_home[["THA"]][groups[[a]], groups[[b]],
                                                        drop = FALSE]))

cat("Prem 2021 home matrix for Thailand:\n")
print(round(C_prem, 3))

# ── Omicron Gamma infectiousness profile ─────────────────────────────────────
mu_GT  <- 3.3; sd_GT  <- 1.5
shape  <- mu_GT^2 / sd_GT^2
rate   <- mu_GT / sd_GT^2
D_max  <- 14
f_raw  <- diff(pgamma(0:D_max, shape = shape, rate = rate))
f_prof <- f_raw / sum(f_raw)
```

------------------------------------------------------------------------

# Helper: run one scenario with caching

```{r}
run_scenario <- function(id, hh_list, mics_priors, C_mat,
                          n_iter = N_ITER, n_burnin = N_BURNIN, n_thin = N_THIN) {
  cache <- sprintf("results/mcmc_%s.rds", id)
  if (file.exists(cache)) {
    cat(sprintf("[%s] loading cache\n", id))
    return(readRDS(cache))
  }
  cat(sprintf("[%s] running MCMC (%d iterations)...\n", id, n_iter))
  res <- run_mcmc(hh_list, mics_priors, C_mat,
                  n_iter = n_iter, n_burnin = n_burnin, n_thin = n_thin,
                  f_prof = f_prof)
  saveRDS(res, cache)
  res
}

# Summarise a result list into a one-row data frame.
summarise_scenario <- function(id, label, res) {
  sar <- (n_obs_infected + res$E_miss_infected) / (n_obs_contacts + n_total_missing)
  data.frame(
    id        = id,
    label     = label,
    q_med     = median(res$q_chain),
    q_lo      = quantile(res$q_chain, 0.025),
    q_hi      = quantile(res$q_chain, 0.975),
    sar_med   = median(sar) * 100,
    sar_lo    = quantile(sar, 0.025) * 100,
    sar_hi    = quantile(sar, 0.975) * 100,
    stringsAsFactors = FALSE
  )
}
```

------------------------------------------------------------------------

# Scenario 1 — Primary (τ=2, Prem matrix)

Load from the cache written by `chain_binomial_mcmc.qmd`. If it does not exist yet,
run the primary analysis here.

```{r}
hh_primary    <- build_hh_list(dataset, tau = 2)
mics_primary  <- compute_mics_priors(hh_primary, comp_table, dataset)
res_primary   <- run_scenario("primary", hh_primary, mics_primary, C_prem)
```

------------------------------------------------------------------------

# Scenario 2 — τ sensitivity (τ=1 and τ=3)

Tests sensitivity to the assumed index pre-enrollment infectious period.
τ=1: faster detection (same day as symptom onset).
τ=3: upper limit allowed by the 48-hour enrollment criterion.

```{r}
hh_tau1   <- build_hh_list(dataset, tau = 1)
hh_tau3   <- build_hh_list(dataset, tau = 3)
mics_tau1 <- compute_mics_priors(hh_tau1, comp_table, dataset)
mics_tau3 <- compute_mics_priors(hh_tau3, comp_table, dataset)

res_tau1 <- run_scenario("tau1", hh_tau1, mics_tau1, C_prem)
res_tau3 <- run_scenario("tau3", hh_tau3, mics_tau3, C_prem)
```

------------------------------------------------------------------------

# Scenario 3 — Age-naive model (no contact matrix)

Replaces C_prem with a constant-1 matrix so every age pair has the same weight.
Tests whether the age structure from Prem et al. materially affects the corrected SAR.
If results are similar to the primary analysis, the contact matrix assumption is not
driving conclusions.

```{r}
C_naive <- matrix(1, 3, 3, dimnames = list(AGE_CATS, AGE_CATS))

res_naive <- run_scenario("naive", hh_primary, mics_primary, C_naive)
```

------------------------------------------------------------------------

# Scenario 4 — Density-dependent contacts (γ=0.5 and γ=1.0)

Implements the correction suggested by Goeyvaerts et al. (2018): within-household
contact density decreases with household size.  The effective contact matrix for
household h becomes C_prem / n_h^γ.

γ=0 is the current model (density-independent, no scaling).
γ=0.5 is an intermediate correction.
γ=1.0 is fully frequency-dependent (equivalent to Cauchemez's β/n formulation).

This requires building per-household scaled matrices inside the MCMC.  The cleanest
implementation wraps the standard run_mcmc with a pre-scaled hh_list: for each
household, divide C_mat by n_h^γ and pass the result as C_mat.  Because run_mcmc
takes a single C_mat, a small wrapper creates household-specific lists.

```{r}
# Helper: run the model with per-household contact matrix scaling.
# For each household, the effective contact matrix is C_mat / n_h^gamma.
# This is a lightweight wrapper: it creates n_h-specific C_mat copies and
# modifies the household list so each household carries its own scaled matrix.
# NOTE: requires a modified likelihood that accepts per-household C_mat.
# TODO: implement run_mcmc_density() that accepts a named list C_mat_list
#       keyed by hhid, or add gamma as a direct argument to run_mcmc.

# Placeholder — replace with the density-aware MCMC call when implemented.
# res_gamma05 <- run_scenario_density("gamma05", hh_primary, mics_primary,
#                                      C_prem, gamma = 0.5)
# res_gamma10 <- run_scenario_density("gamma10", hh_primary, mics_primary,
#                                      C_prem, gamma = 1.0)

cat("Density-dependent scenarios not yet implemented — see TODO above.\n")
```

------------------------------------------------------------------------

# Results: comparison table

```{r}
results <- bind_rows(
  summarise_scenario("primary", "Primary (τ=2, Prem)",    res_primary),
  summarise_scenario("tau1",    "τ = 1",                  res_tau1),
  summarise_scenario("tau3",    "τ = 3",                  res_tau3),
  summarise_scenario("naive",   "Age-naive (C = 1)",      res_naive)
  # Add density-dependent rows here once implemented:
  # summarise_scenario("gamma05", "Density γ = 0.5",      res_gamma05),
  # summarise_scenario("gamma10", "Density γ = 1.0",      res_gamma10)
)

knitr::kable(
  results %>%
    select(label, q_med, q_lo, q_hi, sar_med, sar_lo, sar_hi) %>%
    mutate(across(where(is.numeric), \(x) round(x, 3))),
  col.names = c("Scenario", "q median", "q 2.5%", "q 97.5%",
                "SAR % median", "SAR % 2.5%", "SAR % 97.5%"),
  caption = "Posterior summaries across sensitivity scenarios."
)
```

------------------------------------------------------------------------

# Results: forest plot

```{r fig.width=8, fig.height=5}
results$label <- factor(results$label, levels = rev(results$label))

ggplot(results, aes(x = sar_med, y = label)) +
  geom_point(size = 3, colour = "#457B9D") +
  geom_errorbarh(aes(xmin = sar_lo, xmax = sar_hi),
                 height = 0.25, colour = "#457B9D") +
  geom_vline(xintercept = results$sar_med[results$id == "primary"],
             linetype = "dashed", colour = "#E63946", linewidth = 0.6) +
  labs(title    = "Corrected SAR — sensitivity analysis",
       subtitle = "Dashed line = primary estimate",
       x = "Corrected individual-level SAR (%)",
       y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

```{r fig.width=8, fig.height=5}
ggplot(results, aes(x = q_med, y = label)) +
  geom_point(size = 3, colour = "#2a9d8f") +
  geom_errorbarh(aes(xmin = q_lo, xmax = q_hi),
                 height = 0.25, colour = "#2a9d8f") +
  geom_vline(xintercept = results$q_med[results$id == "primary"],
             linetype = "dashed", colour = "#E63946", linewidth = 0.6) +
  labs(title    = "Posterior of q — sensitivity analysis",
       subtitle = "Dashed line = primary estimate",
       x = "q (per-contact-per-day transmission probability)",
       y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```
